Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality

### 2021-11-22 Code/Manuscript-SMOTE-outbreak.Rmd
including multiple ways for determining presence

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

@@

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

@@

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

@@

```{r set up RF helper model}
rf.mod <- function(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO"), datBB.c=df[df$focalhost=="Any",]){
    splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test   and training datasets stratified by the response class
  datz.train <- training(splitit) # create object with training data
  datz.test <- testing(splitit) # create object with testing data
  datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to   allow for cross validation (again stratified by the response class) for optimization of   model hyperparameters

@@

  # partial dependence
  pdale <- NULL
  for(z in vars.final){
    pdz <- partial(ranger_obj, train=datz.train, pred.var=z,   which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T,   trim.outliers=T)
    pdz$var <- z
    colnames(pdz) <- c("value","yhat","var")
    pdz$method <- "PD"

@@

    pdale <- rbind(pdale, pdz)
    pdale <- rbind(pdale, alez)
  }
  write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
}
```

```{r fit MPB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit SB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentSB", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit WBBB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentWBBB", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit MPB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentMPBo", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit SB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentSBo", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit WBBB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentWBBBo", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
```

@@

@@

```{r BB occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

@@

ggpallpres <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-hostpresence.jpg"), plot=ggpallpres, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affectedo) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

@@

ggpallout <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hostspresence.jpg"), plot=ggpallout, width=fig.width1,height=fig.width1,units="in")
```

```{r BB occurrence by suitablehosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affected) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

@@

ggpsuitablepres <- ggplot(resp.suitablehosts, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-suitablehosts.jpg"), plot=ggpsuitablepres, width=fig.width1,height=fig.width1,units="in")

@@

```{r BB outbreak occurrence by suitable hosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affectedo) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

@@

ggpsuitableoutbreak<- ggplot(resp.suitablehostso, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-suitablehosts.jpg"), plot=ggpsuitableoutbreak, width=fig.width1,height=fig.width1,units="in")
```

```{r BB occurrence by hosts25}
## 
res <- df[!df$hosts25=="none",] %>% count(hosts25, affected) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
x <-  df[!df$hosts25=="none",] %>% count(hosts25)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "b", "c", "c", "c", "bc")

resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.hosts25 <- resp

ggphost25pres <- ggplot(resp.hosts25, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-hosts25.jpg"), plot=ggphost25pres, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak occurrence by hosts25}
## Hosts 25
res <- df[!df$hosts25=="none",] %>% count(hosts25, affectedo) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
x <- df[!df$hosts25=="none",] %>% count(hosts25)

resp <- res[res$affectedo=="affected",]
resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "abc", "b", "b", "c", "c", "c")

resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.hosts25o <- resp

ggphost25outbreak <- ggplot(resp.hosts25o, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hosts25.jpg"), plot=ggphost25outbreak, width=fig.width1,height=fig.width1,units="in")
```

```{r BB severity by hosts, eval=F}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]

@@

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "ab", "bc", "c", "cd", "d", "d")
res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggpallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 + 4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentpresence-hostpresence.jpg"), plot=ggpallseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak severity by hosts, eval=F}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affectedo=="affected",]

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "bc", "bc", "cd", "cd", "d")
res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggoallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey") + geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts") + theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-hostpresence.jpg"), plot=ggoallseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

@@

res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA),n=n())

res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "b", "bc", "b", "c", "bc")

ggpsuitablehostsseverity <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +2, fill=Group),size = 3)+ ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentpresence-suitbalehosts.jpg"), plot=ggpsuitablehostsseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affectedo=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

>>>>

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

@@

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

df$nhosts25F <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

@@

@@

```{r fit MPB RF model, eval=F,fig.show='hide'}
response <- "agentMPB"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit) # create object with training data
datz.test <- testing(splitit) # create object with testing data
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

@@

# partial dependence
pdale <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pdz$method <- "PD"

@@

  pdale <- rbind(pdale, pdz)
  pdale <- rbind(pdale, alez)
}
write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit SB RF model, eval=F}
response <- "agentSB"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# partial dependence
pdale <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pdz$method <- "PD"

  pdale <- rbind(pdale, pdz)
  pdale <- rbind(pdale, alez)
}
write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit WBBB RF model, eval=F}
response <- "agentWBBB"
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
datBB.c <- df[df$focalhost=="Any",]

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# set sample weights
wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# partial dependence
pdale <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pdz$method <- "PD"

  pdale <- rbind(pdale, pdz)
  pdale <- rbind(pdale, alez)
}
write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit MPB RF model outbreak, eval=F}
response <- "agentMPBo"
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit) # create object with training data
datz.test <- testing(splitit) # create object with testing data
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit SB RF model outbreak, eval=F}
response <- "agentSBo"
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit) # create object with training data
datz.test <- testing(splitit) # create object with testing data
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

```{r fit WBBB RF model outbreak, eval=F}
response <- "agentWBBBo"
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") # predictor variables
datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)

splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
datz.train <- training(splitit) # create object with training data
datz.test <- testing(splitit) # create object with testing data
datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
```

@@

ggplot(hostz, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position = position_dodge2(width = 0.9, preserve = "single"))+theme(legend.position="bottom")+xlab("")+ylab("number of plots")

@@

```{r occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

@@

ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
```

```{r outbreak occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affectedo) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

@@

ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
```

```{r occurrence by suitablehosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affected) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

@@

ggplot(resp.suitablehosts, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

@@

```{r outbreak occurrence by suitable hosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affectedo) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

@@

ggplot(resp.suitablehostso, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
```

```{r severity by hosts, eval=F}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred$hosts <- factor(dfpred$hosts, levels=levels(dfpred$hosts),  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

@@

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA))

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "a", "b", "bc", "c", "c", "c")

ggsi <- ggplot(dfpred, aes(x=hosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/FigureSeverityby-AllHosts-outbreak.jpg"), plot=ggsi, width=fig.width1,height=fig.width1,units="in")
```

```{r severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

@@

res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA))

res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "b", "c", "bc", "c", "c")

ggss <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/FigureSeverityby-Suitablehosts-outbreak.jpg"), plot=ggss, width=fig.width1,height=fig.width1,units="in")
```

```{r severity combo, eval=F}
ggs <- ggsi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggss + plot_layout(ncol=1)

ggsave( here("Results/Figures/FigureSeverityby-Hosts-outbreak.jpg"), plot=ggs,  width=fig.width1, height=fig.width1*2,units="in")
```

```{r determine co-occurrence of agents by hosts, eval=F}
df.all <- df[df$focalagent=="Any",]

obsexp.dat <- as.data.frame(df.all  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
obsexp.dat$n.exp <- NA
obsexp.dat$freq.exp <- NA
